Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-03-29 doc/thesis/conclusion.tex
Writing stuff

\section{Extensions}

@@

It is important to note that our kernel-ABC method takes a transmission tree as
input, rather than a viral phylogeny. Thus, we have left the estimation of a
transmission tree up to the user. There were two reasons for this choice. First
and foremost, we wished again to avoid extra complexity and keep the number of
estimated parameters small. In theory, it is possible to incorporate the
process by which a viral phylogeny is generated along with a transmission tree
into our method, for example by simulating within-host dynamics. Although this
may be an avenue for future extension, we felt that it would obscure the
primary purpose of this work, which is to study contact network parameters.
Second, there are a number of different methods available for inferring
transmission trees~\autocite{didelot2014bayesian, yima2012unravelling,
jombart2011reconstructing, cottam2008integrating, poon2015phylodynamic}, some
of which incorporate geographic and/or epidemiological data not accomodated by
our method. We therefore felt it would be best to allow researchers to use
their own preferred tree building method.

>>>>

Contact network structure has a substantial impact on epidemic trajectory
\autocite{barthelemy2005dynamical, volz2007susceptible, volz2008sir}, and a few
methods have been developed to estimate network parameters from epidemiological
data \autocite{volz2007susceptible, britton2002bayesian,
groendyke2011bayesian}. It is known that contact network structure can have a
substantial impact on transmission tree shape \autocite{o2010contact,
robinson2013dynamics, leventhal2012inferring, colijn2014phylogenetic,
villandre2016assessment}.

Our work had three main aims. First, we developed a method to estimate contact
network model parameters from phylogenetic data. This method widens the field
of models which may be investigated in a phylodynamic context. Second, we
investigated the parameters of the \gls{BA} model. We determined through
simulation studies that the preferential attachment power $\alpha$ and number
of infected nodes $I$ had a substantial impact on transmission tree shape, and
could be estimated using our method.

@@

It is important to note that our kernel-ABC method takes a transmission tree as
input, rather than a viral phylogeny. Thus, we have left the estimation of a
transmission tree up to the user. There were two reasons for this choice. First
and foremost, we wished again to avoid extra complexity and keep the number of
estimated parameters small. In theory, it is possible to incorporate the
process by which a viral phylogeny is generated along with a transmission tree
into our method, for example by simulating within-host dynamics. Although this
may be an avenue for future extension, we felt that it would obscure the
primary purpose of this work, which is to study contact network parameters.
Second, there are a number of different methods available for inferring
transmission trees~\autocite{didelot2014bayesian, ypma2012unravelling,
jombart2011reconstructing, cottam2008integrating, poon2015phylodynamic}, some
of which incorporate geographic and/or epidemiological data not accommodated by
our method. We therefore felt it would be best to allow researchers to use
their own preferred tree building method.

### 2016-03-29 doc/thesis/intro.tex
Writing stuff

Due to the internal nodes, each infected individual in an epidemic may appear
in the transmission tree more than once. This is different from the
transmission \emph{network}, in which each infected individual appears exactly
once, and edges are in one-to-one correspondence with
transmissions~\autocite{welch2011statistical, keeling2005networks}.
Transmission networks are discussed further in \cref{sec:contactnet}, and the
distinction between the two objects is illustrated in \cref{fig:contactnet}.
However, since transmission networks generally have no cycles (unless
re-infection occurs), they are trees in the graph theoretical sense, and hence
are sometimes also referred to as transmission
trees~\autocite[\eg][]{kenah2015algorithms}. In this work, we reserve the term
``transmission tree'' for the objects depicted on the right side of
\cref{fig:contactnet}, following~\autocite[\eg][]{stadler2013uncovering}. The
term ``transmission network'' is taken to mean the subgraph of the contact
network along which transmissions occured,
following~\autocite{welch2011statistical, keeling2005networks}.

@@

@@

From a public health perspective, knowledge of contact networks has the
potential to be extremely useful. On a population level, network structure can
dramatically affect the speed and pattern of epidemic
spread~\autocite[\eg][]{barthelemy2005dynamical, volz2008sir}. For example,
epidemics are expected to spread more rapidly in networks having the ``small
world'' property, where the average path length between two nodes in the
network is relatively low~\autocite{watts1998collective}. Some sexually
transmitted infections would not be expected to survive in a homogeneously
mixed population, but their long-term persistance can be explained by contact
heterogeneity~\autocite{anderson1992infectious, pastor2001epidemic}. Hence, the
contact network can provide an idea of what to expect as an epidemic unfolds.
In terms of actionable information, vaccination strategies which would
irradicate an epidemic in a random network might not work if the network is
scale-free~\autocite[][see \cref{subsec:pa}]{keeling2005networks}. On a local
level, contact networks can be informative about the groups or individuals who
are at highest risk of acquiring or transmitting infection, and would therefore
benefit most from public health interventions~\autocite{wang2015targeting,
little2014using}.

@@

An alternative approach has been the analysis of other networks, which can be
estimated with phylogenetic methods from viral sequence data. Some work focuses
on the \defn{phylogenetic network}, in which two nodes are connected if the
genetic distance between their viral sequences is below some threshold.
Primarily, this work has focused on the detection of \defn{phylogenetic
clusters}, which are groups of individuals whose viral sequences are
significantly more similar to each other's than to the general population's.
The phylogenetic network is informative about ``hotspots'' of transmission and
can be used to identify demographic groups to whom targeted intervetions are
likely to have the greatest effect~\autocite{poon2014impact}.  However, this
network may show little to no agreement with a contact data obtained through
epidemiological methods~\autocite{yirrell1998molecular, resik2007limitations,
robinson2013dynamics}, and therefore may be a poor proxy for the contact
network. Other studies~\autocite{brown2011transmission} have investigated the
\defn{transmission network}, which is the subgraph of the contact network
consisting of infected nodes and the edges which led to their
infections~\autocite{welch2011statistical} (\cref{fig:contactnet}, left). It is
possible to estimate the transmission network phylogenetically, although the
methods required for doing so are more sophisticated than for estimating the
phylogenetic network~\autocite{brown2011transmission}. These studies again
mostly focusing on clustering, and also on degree distributions.

@@

A number of studies have made progress in quantifying the relationship between
contact networks and transmission trees. \textcite{o2010contact} simulated
epidemics over networks with four types of degree distribution. They then
estimated the Bayesian skyride~\autocite{minin2008smooth} population size
trajectory in two ways: from the phylogeny, using \gls{MCMC}; and from the
incidence and prevalence trajectories, using the method developed by 
\textcite{volz2009phylodynamics}. They found that the concordance between
the two skyrides, as well as the relationship between the skyride and
prevalence curve, was qualitatively different for each degree distribution.
\citeauthor{leventhal2012inferring} simulated transmission trees over \gls{ER},
\gls{WS}, \gls{BA}, and full networks with fixed number of nodes and mean
degree. They calculated Sackin's index of the simulated trees while varying the
epidemic, network, and sampling parameters, and found that the relationship
between these parameters and Sackin's index varies considerably among the
different network models. In summary, studies in this group have demonstrated
that network structure profoundly influences tree shape, but have not attempted
to quantitatively infer network parameters from observed trees.

@@

\subsection{Model fitting}

A \defn{mathematical model} is a formal description of a hypothesized
relationship between some observed data, $\vec{x} = \set{x_1, \ldots, x_n}$,
and outcomes, $\vec{y} = \set{y_1, \ldots, y_n}$. A \defn{parametric} model
defines a family of possible relationships between data and outcomes, indexed
by one or more numeric parameters $\theta$. A \defn{statistical} model
describes the relationship between data and outcomes in terms of probabilities.
Statistical models define, either explicitly or implicitly, the probability of
observing $\vec{y}$ given $\vec{x}$ and, if the model is parameteric, $\theta$.
In this context, the observed outcomes are taken to be realizations of random
variables $\vec{Y} = \set{Y_1, \ldots, Y_n}$. Note that it is entirely possible
to have no data $\vec{x}$, only observed outcomes $\vec{y}$. In this case, a
model would describe the process by which $\vec{y}$ is generated.

@@

\Gls{ML} estimation makes use only of the data and outcomes to estimate the
model parameters $\theta$. However, it is frequently the case that the
investigator has some additional information or \defn{prior belief} about what
$\theta$ are likely to be. For example, in the linear regression case, the
intstrument used to measure the outcomes may have a well-known margin of error,
or the sign of the slope may be obvious from previous experiments. The Bayesian
approach to model fitting makes use of this information by codifying the
investigator's beliefs as a \defn{prior distribution} on the parameters,
denoted $\Pr(\theta)$. Instead of considering only the likeilhood, Bayesian
inference focuses on the product of the likelihood and the prior, $\Pr(\vec{y}
\mid \theta) \Pr(\theta)$. Bayes' theorem tells us that this product is related
to the \textit{posterior distribution} on $\theta$,
\[
  \Pr(\theta \mid \vec{y}) = \frac{\Pr(\vec{y} \mid \theta) \Pr(\theta)}
                                  {\Pr(\vec{y})}.
\]

In principle, $\Pr(\vec{y} \mid \theta) \Pr(\theta)$ can be optimized
numerically just like $\L(\theta \mid \vec{y})$, which would also optimize the
posterior distribution. The resulting optimal parameters are called the
\gls{MAP} estimates. However, from a Bayesian perspective, $\theta$ is not a
fixed quantity to be estimated, but rather a random variable with an associated
distribution (the posterior). Therefore, the \gls{MAP} estimate by itself is of
limited value without associated statistics about the posterior distribution,
such as the mean or credible intervals. Unfortunately, to calculate such
statistics, it is necessary to evaluate the normalizing constant
$\Pr(\vec{y})$, which is almost always an intractible integral.

A popular method for circumventing the normalizing constant is the use of
\gls{MCMC} to obtain a sample from the posterior distribution. \Gls{MCMC} works
by defining a Markov chain whose states are indexed by possible model
parameters. The transition probability from state $\theta_1$ to state
$\theta_2$ is taken to be
\[
  \max\left(1, \frac{\Pr(\vec{y} \mid \theta_2) \Pr(\theta_2) q(\theta_2, \theta_1)}
                    {\Pr(\vec{y} \mid \theta_1) \Pr(\theta_2) q(\theta_1, \theta_2)} \right),
\]
where $q(\theta, \theta')$ is a symmetric \defn{proposal distribution} used in
the algorithm to generate the chain. The stationary distribution of this Markov
chain is equal to the posterior distribution on $\theta$. Therefore, if a long
enough random walk is performed on the chain, the distribution of states
visited will be a Monte Carlo approximation of $\Pr(\theta \mid \vec{y})$, from
which we can calculate statistics of interest. Actually performing this random
walk is straightforward and is accomplished via the Metropolis-Hastings
algorithm.

\begin{algorithm}
  \begin{algorithmic}
    \State Draw $\theta$ according to $\Pr(\theta)$
    \Loop
      \State Propose $\theta'$ according to $q(\theta, \theta')$
      \State Accept $\theta \gets \theta'$ with probability
      $\max \left( 1, 
       \dfrac{\Pr(\vec{y} \mid \theta') \Pr(\theta') q(\theta', \theta)}
             {\Pr(\vec{y} \mid \theta\phantom{'}) \Pr(\theta\phantom{'}) q(\theta, \theta')}
       \right)$
    \EndLoop
  \end{algorithmic}
  \label{alg:mh}
\end{algorithm}

\subsection{Overview and motivation for ABC}

Most mathematical models are amenable to fitting via one or both of the
approaches, \gls{ML} or Bayesian inference, discussed above. However, there are
some, particularly in the domain of population genetics, for which calculation
of either the likelihood or the product of the likelihood and the prior may be
infeasible. For example, one or both of these quantities may be expressible
only as an intractible integral. \Gls{ABC} is designed for such cases, where
standard likelihood-based techniques for model fitting cannot be applied.

The first practical application of \gls{ABC} was by
\textcite{tavare1997inferring}, who wanted to calculate the posterior
distribution on the coalescence time $t$ of $n$ sequences with $k$ observed
segregating sites under the coalescent model~\autocite{kingman1982coalescent}.
That is,
\[
  \Pr(t \mid k) \propto \Pr(t) \Pr(k \mid t),
\]
where again we abuse $\Pr$ to refer to a probability density.
\citeauthor{tavare1997inferring} observed that the expected number of
segregating sites depends on the total branch length of the tree, rather than
its height. This enabled them to write an alternative expression for the
product of the likelihood and prior in terms of the total branch length $l$,
\[
  \int_0^\infty \Pr(t, l) \Pr(k \mid t, l) \d l.
\]
While it is unclear how to evaluate this integral, the power of this
formulation is the fact that it is straightforward to \emph{simulate} points
according to $\Pr(t, l)$, the joint probability of $t$ and $l$, by simply
running the (stochastic) coalescent process and observing the values $\bar{t}$
and $\bar{l}$ which result.

%\Gls{ABC} is a statistical method designed to fit complex models, which cannot
%be fit using more conventional methods, to observed
%data~\autocite{marin2012approximate, sunnaker2013approximate,
%beaumont2010approximate}. We shall make this precise below, but to fully
%describe and motivate \gls{ABC}, it is necessary to first explain exactly what
%we mean by ``fitting'' a model. We will illustrate this concept with one of the
%simplest and most ubiquitous models: linear regression.
%
%An approximate Bayesian computation approach to least-squares regression might
%be as follows. We will need to make an additional assumption: that the errors
%$\varepsilon_i$ are normally distributed. That is, our linear model is now of
%the form
%\[
%  y_i = bx_i + \N(0, \sigma),
%\]
%where $\N(0, \sigma)$ indicates a normal distribution with mean zero and
%variance $\sigma$. We do not know $\sigma$ in advance, so this is another
%parameter we will need to estimate. There is nothing special about normal
%distributions - we could have chosen any other distribution we deemed
%appropriate. 
%
%The main idea of \gls{ABC} is to replace the traditional likelihood-based
%criteria for model ``goodness'' with one based on simulated data generated by
%the model. Good models should be able to simulate data which closely resemble
%reality. In \gls{ABC}, we are no longer interested in the posterior
%distribution, but rather in the distribution of model parameters which produce
%data sufficiently ``close'' to the real data. To formalize this, let $d(\cdot,
%\cdot)$ be a distance function on datasets, and $\varepsilon$ be a tolerance
%level. The objective of \gls{ABC} is to recover the distribution\ldots

@@

Continuing with the linear model example, we need to specify a prior $\pi(a, b,
\sigma)$ on the three parameters. We do not have much certain information about
these parameters except that $\sigma$ has to be at least zero, but it seems
reasonable to assume that extreme relationships are fairly rare, and that
positive and negative correlations are equiprobable. Therefore, we will let
$a$, $b$, and $\sigma$ be independent, $a$ and $b$ be normally distributed,
and $\sigma$ be log-normally distributed. That is,
\[
  \pi(a, b, \sigma) = 
  \begin{cases}
    \Pr[\N(0, 1) = a] \cdot \Pr[\N(0, 1) = b] \cdot \Pr[\N(0, 1) = \log\sigma]
     & \sigma > 0 \\
    0 & \sigma \le 0.
  \end{cases}
\]
For the vector of summary statistics, we will use the mean and variance of the
simulated data,
\[
  \eta(\vec{y}) = \left<\E[\vec{y}], \Var[\vec{y}]\right>.
\]
Finally, for the distance function $\rho$ we take the standard Euclidian
distance.

@@

The basis of \gls{SIS} is \gls{IS}, which is a method of estimating summary
statistics of distributions which are known only up to a normalizing constant,
and therefore cannot be sampled from directly. That is, if $\pi$ is such a
distribution and $f$ is any real-valued function, \gls{IS} is concerned with
estimating
\[
  \pi(f) = \int f(x)\pi(x)\d x = \int f(x) \frac{\gamma(x)}{Z} \d x,
\]
where the integral is over the space on which $\pi$ is defined, $\gamma(x)$ is
known, and $Z$ is the unknown normalizing constant, $Z = \int \gamma(x) \d x$.
The posterior distributions of all but the simplest models' parameters fall
into this category. Suppose we have at hand another distribution $\eta$, called
the \defn{importance distribution}, from which we are able to sample. Define
the \defn{importance weight} as the ratio ratio $w(x) = \gamma(x)/\eta(x)$. We
can express the normalizing constant $Z$ in terms of the importance weight and
distribution, $Z = \int w(x) \eta(x) \d x$, and in turn write the original
integral of interest as
\[
  \int f(x) \pi(x) \d x = \frac{\int f(x) \gamma(x) \d x}
                               {\int w(x) \eta(x) \d x}.
\]
If we sample a large number of points from $\eta$, then $\eta(x)$ can be
approximated at each of them by an empirical distribution. Since the remaining
quantities $f$, $\gamma$, and $w$ can all be evaluated pointwise, these are all
the ingredients we need to obtain an estimate of $\pi(f)$. Although this is a
simple and elegant approach, the drawback is that the variance of the estimate
is proportional to the variance of the importance
weights~\autocite{liu2008monte}, which may be quite large if $\eta$ and
$\gamma$ are very different. Therefore, the practical use of \gls{IS} on its
own is limited, since it depends on finding an importance distribution which is
similar to $\pi$, which we usually know very little about \textit{a priori}.

>>>>

Due to the internal nodes, each infected individual in an epidemic may appear
in the transmission tree more than once. This is different from the
transmission \emph{network}, in which each infected individual appears exactly
once, and edges are in one-to-one correspondence with
transmissions~\autocite{welch2011statistical, keeling2005networks}.
Transmission networks are discussed further in \cref{sec:contactnet}, and the
distinction between the two objects is illustrated in \cref{fig:contactnet}.
However, since transmission networks generally have no cycles (unless
re-infection occurs), they are trees in the graph theoretical sense, and hence
are sometimes also referred to as transmission
trees~\autocite[\eg][]{kenah2015algorithms}. In this work, we reserve the term
``transmission tree'' for the objects depicted on the right side of
\cref{fig:contactnet}, following~\autocite[\eg][]{stadler2013uncovering}. The
term ``transmission network'' is taken to mean the subgraph of the contact
network along which transmissions occurred,
following~\autocite{welch2011statistical, keeling2005networks}.

@@

\subsection{Applications of phylodynamics}
\label{subsec:phyloapp}

Phylodynamic methods have been used to investigate epidemiological parameters
such as transmission rate, recovery rate, and basic reproductive
number~\autocite{pybus2009evolutionary, volz2013viral}. These studies make
inferences about epidemiological processes from the genetic diversity of virus
populations, which is usually represented in the form of a phylogeny. The
majority of these employ a Bayesian \gls{MCMC} approach to infer parameters of
an epidemiological model whose likelihood can be calculated, most often some
variation of the birth-death~\autocite{kendall1948generalized} or
coalescent~\autocite{kingman1982coalescent} models.
\textcite{stadler2011estimating} develop a formula for the likelihood of a
phylogeny with heterochronous tips under the birth-death model, which has been
used to estimate the basic reproductive number of several viral
epidemics~\autocite{stadler2011estimating}. However, the birth-death model is
cannot tell us anything about population structure, as it assumes that every
individual becomes infected at the same rate. \textcite{volz2012complex} writes
down the likelihood of a heterochronous phylogeny under a coalescent model with
arbitrarily complex population dynamics. This opens the door to more complex
inferences about population structure, as the population can be partitioned
into compartments with different transmission and recovery rates, but still
assumes that each compartment is homogeneously mixed. In other words, the
coalescent model can tell us about the \emph{global} structure of a population,
such as whether there exists a high-risk subgroup, but not about the
\emph{local} structure, such as the average number of contacts each individual
has.

@@

From a public health perspective, knowledge of contact networks has the
potential to be extremely useful. On a population level, network structure can
dramatically affect the speed and pattern of epidemic
spread~\autocite[\eg][]{barthelemy2005dynamical, volz2008sir}. For example,
epidemics are expected to spread more rapidly in networks having the ``small
world'' property, where the average path length between two nodes in the
network is relatively low~\autocite{watts1998collective}. Some sexually
transmitted infections would not be expected to survive in a homogeneously
mixed population, but their long-term persistence can be explained by contact
heterogeneity~\autocite{anderson1992infectious, pastor2001epidemic}. Hence, the
contact network can provide an idea of what to expect as an epidemic unfolds.
In terms of actionable information, vaccination strategies which would
eradicate an epidemic in a random network might not work if the network is
scale-free~\autocite[][see \cref{subsec:pa}]{keeling2005networks}. On a local
level, contact networks can be informative about the groups or individuals who
are at highest risk of acquiring or transmitting infection, and would therefore
benefit most from public health interventions~\autocite{wang2015targeting,
little2014using}.

@@

An alternative approach has been the analysis of other networks, which can be
estimated with phylogenetic methods from viral sequence data. Some work focuses
on the \defn{phylogenetic network}, in which two nodes are connected if the
genetic distance between their viral sequences is below some threshold.
Primarily, this work has focused on the detection of \defn{phylogenetic
clusters}, which are groups of individuals whose viral sequences are
significantly more similar to each other's than to the general population's.
The phylogenetic network is informative about ``hotspots'' of transmission and
can be used to identify demographic groups to whom targeted interventions are
likely to have the greatest effect~\autocite{poon2014impact}.  However, this
network may show little to no agreement with a contact data obtained through
epidemiological methods~\autocite{yirrell1998molecular, resik2007limitations,
robinson2013dynamics}, and therefore may be a poor proxy for the contact
network. Other studies~\autocite{brown2011transmission} have investigated the
\defn{transmission network}, which is the subgraph of the contact network
consisting of infected nodes and the edges which led to their
infections~\autocite{welch2011statistical} (\cref{fig:contactnet}, left). It is
possible to estimate the transmission network phylogenetically, although the
methods required for doing so are more sophisticated than for estimating the
phylogenetic network~\autocite{brown2011transmission}. These studies again
mostly focusing on clustering, and also on degree distributions.

@@

A number of studies have made progress in quantifying the relationship between
contact networks and transmission trees. \textcite{o2010contact} simulated
epidemics over networks with four types of degree distribution. They then
estimated the Bayesian skyride~\autocite{minin2008smooth} population size
trajectory in two ways: from the phylogeny, using \gls{MCMC}; and from the
incidence and prevalence trajectories, using the method developed by 
\textcite{volz2009phylodynamics}. They found that the concordance between
the two skyrides, as well as the relationship between the skyride and
prevalence curve, was qualitatively different for each degree distribution.
\textcite{leventhal2012inferring} investigated the relationship between
transmission tree imbalance and several epidemic parameters under four contact
network models, and found that these relationships varied considerably
depending on which model was being considered. \textcite{welch2011network}
simulated transmission trees over networks with varying degrees of community
structure. They found that transmission trees simulated under networks with low
clustering could not generally be distinguished from those simulated under
highly clustered networks, and concluded that contact network clusters do not
affect transmission tree shape. However, more recently,
\textcite{villandre2016assessment} investigated the correspondence between
contact network clusters and transmission tree clusters, and found a moderate
correspondence between the two.

In summary, studies in this group have demonstrated that network structure
profoundly influences tree shape, but have not attempted to quantitatively
infer network parameters from observed trees.

@@

\subsection{Model fitting}
\label{subsec:mfit}

A \defn{mathematical model} is a formal description of a hypothesized
relationship between some observed data, $\vec{x} = \set{x_1, \ldots, x_n}$,
and outcomes, $\vec{y} = \set{y_1, \ldots, y_n}$. A \defn{parametric} model
defines a family of possible relationships between data and outcomes, indexed
by one or more numeric parameters $\theta$. A \defn{statistical} model
describes the relationship between data and outcomes in terms of probabilities.
Statistical models define, either explicitly or implicitly, the probability of
observing $\vec{y}$ given $\vec{x}$ and, if the model is parametric, $\theta$.
In this context, the observed outcomes are taken to be realizations of random
variables $\vec{Y} = \set{Y_1, \ldots, Y_n}$. Note that it is entirely possible
to have no data $\vec{x}$, only observed outcomes $\vec{y}$. In this case, a
model would describe the process by which $\vec{y}$ is generated.

@@

\Gls{ML} estimation makes use only of the data and outcomes to estimate the
model parameters $\theta$. However, it is frequently the case that the
investigator has some additional information or \defn{prior belief} about what
$\theta$ are likely to be. For example, in the linear regression case, the
instrument used to measure the outcomes may have a well-known margin of error,
or the sign of the slope may be obvious from previous experiments. The Bayesian
approach to model fitting makes use of this information by codifying the
investigator's beliefs as a \defn{prior distribution} on the parameters,
denoted $\Pr(\theta)$. Instead of considering only the likelihood, Bayesian
inference focuses on the product of the likelihood and the prior, $\Pr(\vec{y}
\mid \theta) \Pr(\theta)$. Bayes' theorem tells us that this product is related
to the \textit{posterior distribution} on $\theta$,
\[
  \Pr(\theta \mid \vec{y}) = \frac{\Pr(\vec{y} \mid \theta) \Pr(\theta)}
                                  {\Pr(\vec{y})}.
\]

In principle, $\Pr(\vec{y} \mid \theta) \Pr(\theta)$ can be optimized
numerically just like $\L(\theta \mid \vec{y})$, which would also optimize the
posterior distribution. The resulting optimal parameters are called the
\gls{MAP} estimates. However, from a Bayesian perspective, $\theta$ is not a
fixed quantity to be estimated, but rather a random variable with an associated
distribution (the posterior). Therefore, the \gls{MAP} estimate by itself is of
limited value without associated statistics about the posterior distribution,
such as the mean or credible intervals. Unfortunately, to calculate such
statistics, it is necessary to evaluate the normalizing constant
$\Pr(\vec{y})$, which is almost always an intractable integral.

A popular method for circumventing the normalizing constant is the use of
\gls{MCMC} to obtain a sample from the posterior distribution. \Gls{MCMC} works
by defining a Markov chain whose states are indexed by possible model
parameters. The transition probability from state $\theta_1$ to state
$\theta_2$ is taken to be
\[
  \max\left(1, \frac{\Pr(\vec{y} \mid \theta_2) \Pr(\theta_2) q(\theta_2, \theta_1)}
                    {\Pr(\vec{y} \mid \theta_1) \Pr(\theta_2) q(\theta_1, \theta_2)} \right),
\]
where $q(\theta, \theta')$ is a symmetric \defn{proposal distribution} used in
the algorithm to generate the chain. The stationary distribution of this Markov
chain is equal to the posterior distribution on $\theta$. Therefore, if a long
enough random walk is performed on the chain, the distribution of states
visited will be a Monte Carlo approximation of $\Pr(\theta \mid \vec{y})$, from
which we can calculate statistics of interest. Actually performing this random
walk is straightforward and is accomplished via the Metropolis-Hastings
algorithm (\cref{alg:mh}).

\begin{algorithm}
  \label{alg:mh}
  \caption{Metropolis-Hastings algorithm for Markov chain Monte Carlo.}
  \begin{algorithmic}
    \State Draw $\theta$ according to $\Pr(\theta)$
    \Loop
      \State Propose $\theta'$ according to $q(\theta, \theta')$
      \State Accept $\theta \gets \theta'$ with probability
      $\max \left( 1, 
       \dfrac{\Pr(\vec{y} \mid \theta') \Pr(\theta') q(\theta', \theta)}
             {\Pr(\vec{y} \mid \theta\phantom{'}) \Pr(\theta\phantom{'}) q(\theta, \theta')}
       \right)$
    \EndLoop
  \end{algorithmic}
  \label{alg:mh}
\end{algorithm}

\subsection{Overview and motivation for ABC}
\label{subsec:abcoverview}

Most mathematical models are amenable to fitting via one or both of the
approaches, \gls{ML} or Bayesian inference, discussed above. However, there are
some, particularly in the domain of population genetics, for which calculation
of either the likelihood or the product of the likelihood and the prior may be
infeasible. For example, one or both of these quantities may be expressible
only as an intractable integral. \Gls{ABC} is designed for such cases, where
standard likelihood-based techniques for model fitting cannot be applied. Such
models are particularly prevalent in population
genetics~\autocite{beaumont2002approximate, beaumont2010approximate}. We will
begin this subsection by describing \gls{ABC} in general terms. Then, we will
use linear regression as a toy problem to demonstrate how \gls{ABC} could be
applied. Finally, we will give an example of a model which is impractical to
fit using likelihood-based methods, and show how its parameters were estimated
using \gls{ABC}.

Ordinarily, Bayesian inference targets the posterior distribution.

%traditional likelihood criterion
%for model ``goodness'' with one based on simulated data generated by the model.
%Good models should be able to simulate data which closely resemble reality. In
%\gls{ABC}, we are no longer interested in the posterior distribution, but
%rather in the distribution of model parameters which produce data sufficiently
%``close'' to the real data. To formalize this, let $d(\cdot, \cdot)$ be a
%distance function on datasets, and $\varepsilon$ be a tolerance level.

Of course, linear models can be fit more easily using likelihood-based methods, 
since the likelihood $\L(\theta \mid \vec{y})$ can be easily calculated (see
\cref{subsec:fit}). One model where this is not the case is Kingman's
coalescent~\autocite{kingman1982coalescent}.

As an example of such a model, we will review the first practical application
of \gls{ABC} by \textcite{tavare1997inferring}. The aim of the investigation
was to calculate the posterior distribution on the coalescence time $t$ of some
observed sequences $D$ under the coalescent model
\autocite{kingman1982coalescent} with the infinite-sites assumption
\autocite{watterson1975number}. Their estimation method relied on two key
observations. First, the number $k$ of \defn{segregating sites} in the data is
informative about the coalescence time while being much easier to work with
than the data itself. A segregating site is a position which is polymorphic in
the observed sequences. Using this observation, the authors replaced their
posterior distribution of interest, $\Pr(t \mid D)$, by
\[
  \Pr(t \mid k) \propto \Pr(t) \Pr(k \mid t),
\]
where again we abuse $\Pr$ to refer to a probability density. Second,
the authors observed that the expected number of segregating sites depends on
the total branch length $l$ of the tree, rather than its height. This enabled
them to write an alternative expression for the product of the likelihood and
prior in terms of $l$,
\[
  \Pr(t) \Pr(k \mid t) = \int_0^\infty \Pr(t, l) \Pr(k \mid l) \d l.
\]
While this integral may be difficult to evaluate analytically, the power of
this formulation is the fact that it is straightforward to \emph{simulate}
values of $t$ and $l$ according to $\Pr(t, l)$, their joint probability under
the coalescent model. One simply runs the (stochastic) coalescent process and
observing the values $\hat{t}$ and $\hat{l}$ which result. It is
straightforward to calculate $\Pr(k \mid \hat{l})$, the probability
that $k$ segregating sites were observed in a tree with total branch length
$\hat{l}$, using the Poisson distribution. Hence if $\hat{l}$ and $\hat{t}$,
having been generated according to $\Pr(t, l)$, are sampled with probability
proportional to $\Pr(k \mid \hat{l})$, they represent an unbiased sample from
the above integral. Taking a large number of such samples allowed the authors
to obtain a Monte Carlo estimate of $\Pr(t \mid k)$.

%\Gls{ABC} is a statistical method designed to fit complex models, which cannot
%be fit using more conventional methods, to observed
%data~\autocite{marin2012approximate, sunnaker2013approximate,
%beaumont2010approximate}. We shall make this precise below, but to fully
%describe and motivate \gls{ABC}, it is necessary to first explain exactly what
%we mean by ``fitting'' a model. We will illustrate this concept with one of the
%simplest and most ubiquitous models: linear regression.
%
%An approximate Bayesian computation approach to least-squares regression might
%be as follows. We will need to make an additional assumption: that the errors
%$\varepsilon_i$ are normally distributed. That is, our linear model is now of
%the form
%\[
%  y_i = bx_i + \N(0, \sigma),
%\]
%where $\N(0, \sigma)$ indicates a normal distribution with mean zero and
%variance $\sigma$. We do not know $\sigma$ in advance, so this is another
%parameter we will need to estimate. There is nothing special about normal
%distributions - we could have chosen any other distribution we deemed
%appropriate. 
%

@@

Continuing with the linear model example, we need to specify a prior $\pi(a, b,
\sigma)$ on the three parameters. We do not have much certain information about
these parameters except that $\sigma$ has to be at least zero, but it seems
reasonable to assume that extreme relationships are fairly rare, and that
positive and negative correlations are equiprobable. Therefore, we will let
$a$, $b$, and $\sigma$ be independent, $a$ and $b$ be normally distributed,
and $\sigma$ be log-normally distributed. That is,
\[
  \pi(a, b, \sigma) = 
  \begin{cases}
    \Pr[\N(0, 1) = a] \cdot \Pr[\N(0, 1) = b] \cdot \Pr[\N(0, 1) = \log\sigma]
     & \sigma > 0 \\
    0 & \sigma \le 0.
  \end{cases}
\]
For the vector of summary statistics, we will use the mean and variance of the
simulated data,
\[
  \eta(\vec{y}) = \left<\E[\vec{y}], \Var[\vec{y}]\right>.
\]
Finally, for the distance function $\rho$ we take the standard Euclidean
distance.

@@

The basis of \gls{SIS} is \gls{IS}, which is a method of estimating summary
statistics of distributions which are known only up to a normalizing constant,
and therefore cannot be sampled from directly. That is, if $\pi$ is such a
distribution and $f$ is any real-valued function, \gls{IS} is concerned with
estimating
\[
  \pi(f) = \int f(x)\pi(x)\d x = \int f(x) \frac{\gamma(x)}{Z} \d x,
\]
where the integral is over the space on which $\pi$ is defined, $\gamma(x)$ is
known, and $Z$ is the unknown normalizing constant, $Z = \int \gamma(x) \d x$.
The posterior distributions of all but the simplest models' parameters fall
into this category. Suppose we have at hand another distribution $\eta$, called
the \defn{importance distribution}, from which we are able to sample. Define
the \defn{importance weight} as the ratio ratio $w(x) = \gamma(x)/\eta(x)$. We
can express the normalizing constant $Z$ in terms of the importance weight and
distribution, $Z = \int w(x) \eta(x) \d x$, and in turn write the original
integral of interest as
\[
  \int f(x) \pi(x) \d x = \frac{\int f(x) \gamma(x) \d x}
                               {\int w(x) \eta(x) \d x}.
\]
If we sample a large number of points from $\eta$, then $\eta(x)$ can be
approximated at each of them by an empirical distribution. Since the remaining
quantities $f$, $\gamma$, and $w$ can all be evaluated pointwise, these are all
the ingredients we need to obtain an estimate of $\pi(f)$. Although this is a
simple and elegant approach, the drawback is that the variance of the estimate
is proportional to the variance of the importance
weights~\autocite{liu2008monte}, which may be quite large if $\eta$ and
$\gamma$ are very different. Therefore, the practical use of \gls{IS} on its
own is limited, since it depends on finding an importance distribution which is
similar to $\pi$, which we usually know very little about \textit{a priory}.
